---
title: "Apache Parquet-based Mass Spectrometry Data Backend"
author:
- name: Ossama Edbali
package: MsBackendParquet
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Apache Parquet-based Mass Spectrometry Data Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = "asis", message = FALSE}
BiocStyle::markdown()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

The `r Biocpkg("MsBackendParquet")` package provides a backend for
`r Biocpkg("Spectra")` objects that stores mass spectrometry (MS) data
in an on-disk [Apache Parquet](https://parquet.apache.org/) dataset,
accessed through the [Apache Arrow](https://arrow.apache.org/docs/r) R
bindings.

The backend extends `MsBackendCached` from `r Biocpkg("Spectra")` so it
benefits from caching of user-defined spectra variables while keeping
the on-disk dataset read-only.

# Creating a Parquet dataset

A Parquet dataset can be created from raw MS data files (mzML, mzXML,
netCDF) via `createMsBackendParquetDataset()`:

```{r, eval = FALSE}
library(Spectra)
library(MsBackendParquet)

files <- c("sample-1.mzML", "sample-2.mzML")
path  <- tempfile()
createMsBackendParquetDataset(path = path, x = files,
                              partitioning = "dataOrigin",
                              compression = "snappy")
```

The function imports MS data via `MsBackendMzR()` by default and
writes spectra metadata together with the `mz`/`intensity` values as
Parquet `list<double>` columns.

For the rest of the vignette we work with a small in-memory data set
so the document can be built without external files:

```{r build-dataset}
library(Spectra)
library(MsBackendParquet)

sd <- S4Vectors::DataFrame(
    msLevel = c(1L, 1L, 2L),
    rtime = c(1.0, 2.0, 3.0),
    precursorMz = c(NA_real_, NA_real_, 110.5),
    dataOrigin = c("file-a", "file-a", "file-b"))
sd$mz <- IRanges::NumericList(c(100, 110), c(101, 111),
                              c(102, 112), compress = FALSE)
sd$intensity <- IRanges::NumericList(c(10, 20), c(11, 21),
                                     c(12, 22), compress = FALSE)

path <- tempfile()
be <- backendInitialize(MsBackendParquet(), path = path, data = sd)
be
```

# Accessing data

`spectraData()`, `peaksData()`, `mz()`, `intensity()` and friends work
the same way as for any other `MsBackend`:

```{r access}
spectraData(be, c("msLevel", "rtime"))
peaksData(be)[1:2]
mz(be)
```

# Filtering with Arrow predicate push-down

Filter methods are implemented as `dplyr` lazy queries on the open
dataset and benefit from Arrow's predicate push-down:

```{r filtering}
filterMsLevel(be, 1L)
filterRt(be, c(1.5, 2.5))
filterDataOrigin(be, "file-b")
```

# Streaming large mzML files

For very large mzML files, the default conversion path goes through a
`Spectra` object built with `MsBackendMzR()`, which materialises a
whole chunk of peaks in memory before writing it to Parquet. The
`engine = "mzr"` option bypasses `Spectra` entirely and streams
batches of spectra directly from `mzR::openMSfile()` into Apache
Arrow's incremental `ParquetFileWriter`:

```{r, eval = FALSE}
be <- mzMLToParquet("huge-sample.mzML",
                    path = tempfile(),
                    engine = "mzr",
                    batch_size = 1000L)
```

Memory use stays bounded by `batch_size * mean(peaks per spectrum)`,
regardless of total file size. The streaming engine processes files
serially and requires the `mzR` package; it does not honour `BPPARAM`
or `chunksize`.

# Saving and reloading the backend

Because `MsBackendParquet` objects only contain a file system path,
they are fully serialisable and can be sent to parallel workers
without re-opening connections:

```{r save-load}
rds <- tempfile(fileext = ".rds")
saveRDS(be, rds)
be2 <- readRDS(rds)
identical(spectraNames(be), spectraNames(be2))
```

# Session information

```{r si}
sessionInfo()
```
