---
title: "Apache Parquet-based Mass Spectrometry Data Backend"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Apache Parquet-based Mass Spectrometry Data Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{MsBackendParquet}
---

```{r style, echo = FALSE, results = "asis", message = FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendParquet")`<br />
**Authors**: `r packageDescription("MsBackendParquet")[["Author"]] `<br />
**Compiled**: `r date()`

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BiocStyle)
```

# Introduction

The `r Biocpkg("MsBackendParquet")` package provides a backend for
`r Biocpkg("Spectra")` objects that stores mass spectrometry (MS) data
in an on-disk [Apache Parquet](https://parquet.apache.org/) dataset.
Datasets are *written* with the
[Apache Arrow](https://arrow.apache.org/docs/r) R bindings and *queried*
with [DuckDB](https://duckdb.org/), keeping the spectra metadata
together with the `mz`/`intensity` peak values in a columnar format that
supports compression, dictionary encoding and Hive-style partitioning.

To keep access fast, the (small) non-peak spectra variables are held in
a bounded in-memory index while the peak data is read from disk on
demand. As a result, metadata filters cost about as much as with an
in-memory backend, while the bulk peak storage stays out of core.

The backend extends `MsBackendCached` from `r Biocpkg("Spectra")` so it
benefits from caching of user-defined spectra variables while keeping
the on-disk dataset read-only.

# Creating a Parquet dataset

The quickest way to turn raw MS data files (mzML, mzXML, netCDF) into a
Parquet dataset is the `mzMLToParquet()` helper. It validates the input
files, writes the dataset and returns an initialised backend ready to be
wrapped in a `Spectra` object:

```{r, eval = FALSE}
library(Spectra)
library(MsBackendParquet)

files <- c("sample-1.mzML", "sample-2.mzML")
be <- mzMLToParquet(files, path = tempfile(), partitioning = "dataOrigin")
sps <- Spectra(be)
```

For more control (a custom reader backend, chunk sizes, compression
codec, Parquet row-group size, or writing an existing `DataFrame`), call
`createMsBackendParquetDataset()` directly:

```{r, eval = FALSE}
path <- tempfile()
createMsBackendParquetDataset(path = path, x = files,
                              partitioning = "dataOrigin",
                              compression = "snappy")
be <- backendInitialize(MsBackendParquet(), path = path)
```

Both functions import MS data via `MsBackendMzR()` by default and write
spectra metadata together with the `mz`/`intensity` values as Parquet
`list<double>` columns.

For the rest of the vignette we work with a small in-memory data set
so the document can be built without external files:

```{r build-dataset}
library(Spectra)
library(MsBackendParquet)

sd <- S4Vectors::DataFrame(
    msLevel = c(1L, 1L, 2L),
    rtime = c(1.0, 2.0, 3.0),
    precursorMz = c(NA_real_, NA_real_, 110.5),
    dataOrigin = c("file-a", "file-a", "file-b"))
sd$mz <- IRanges::NumericList(c(100, 110), c(101, 111),
                              c(102, 112), compress = FALSE)
sd$intensity <- IRanges::NumericList(c(10, 20), c(11, 21),
                                     c(12, 22), compress = FALSE)

path <- tempfile()
be <- backendInitialize(MsBackendParquet(), path = path, data = sd)
be
```

# Accessing data

`spectraData()`, `peaksData()`, `mz()`, `intensity()` and related work
the same way as for any other `MsBackend`:

```{r access}
spectraData(be, c("msLevel", "rtime"))
peaksData(be)[1:2]
mz(be)
```

# Filtering with DuckDB predicate push-down

Filter methods are translated into DuckDB SQL predicates that are pushed
down to the Parquet scan. DuckDB prunes whole row groups using their
min/max statistics (and skips whole partitions for Hive-partitioned
datasets), so a range or set filter reads only the data it needs. When
the filter column is held in the in-memory metadata index (see below),
the predicate is evaluated in memory instead:

```{r filtering}
filterMsLevel(be, 1L)
filterRt(be, c(1.5, 2.5))
filterDataOrigin(be, "file-b")
filterPrecursorMzRange(be, c(110, 111))
filterPrecursorMzValues(be, 110.5, ppm = 20)
```

Chained filters combine into a single scan, and a subsequent
`peaksData()` call reuses the same predicate to prune the peak data it
reads.

# The in-memory metadata cache

The non-peak spectra variables are typically small compared to the peak
data, so `MsBackendParquet` keeps them in a bounded, read-through
in-memory index. The cache is held at the package level, keyed by
dataset path and shared by every backend over that dataset (so it is
independent of the individual, serialisable backend objects). It is
warmed on `backendInitialize()` and dropped automatically whenever the
dataset on disk is (re-)written. It is entirely separate from the
user-modified variables tracked by `MsBackendCached`, which always take
precedence.

The cache is controlled through three options (shown with their
defaults):

```{r, eval = FALSE}
# Enable ("auto", the default) or disable ("off") the metadata cache.
options(MsBackendParquet.cacheMetadata = "auto")

# Columns held by default (the ones the optimised filters need).
options(MsBackendParquet.cacheColumns =
            c("msLevel", "rtime", "precursorMz", "dataOrigin"))

# Budget in cells (rows x columns); ~5e7 is roughly 400 MB of doubles.
# Requests over budget quietly fall back to reading from disk.
options(MsBackendParquet.cacheMaxCells = 5e7)
```

Two further options tune the underlying DuckDB session:
`MsBackendParquet.threads` (maps to DuckDB's `threads`) and
`MsBackendParquet.memoryLimit` (maps to `memory_limit`). Both default to
DuckDB's own defaults when unset.

# Performance

Storing the peaks as columnar `list<double>` values lets DuckDB skip
whole row groups when a query touches only part of the data, while
projection and predicate push-down keep both the columns and rows read
to a minimum. Combined with the in-memory metadata index, this makes
metadata filters roughly as fast as an in-memory backend while the peak
storage stays on disk.

Across the SpectraQL query suite over 76,020 spectra (10 mzML files),
the Parquet backend is the fastest of the on-disk `Spectra` backends on
5 of 7 queries. For example, MS1 peak retrieval takes about 95 ms versus
roughly 560 ms for the `mzR` backend. The benchmark scripts live in
[`inst/benchmarks/`](https://github.com/computational-metabolomics/MsBackendParquet/tree/main/inst/benchmarks);
see `inst/benchmarks/performance.md` for the full numbers and the design
considerations behind them.

# Streaming large mzML files

For very large mzML files, the default conversion path goes through a
`Spectra` object built with `MsBackendMzR()`, which materialises a
whole chunk of peaks in memory before writing it to Parquet. The
`engine = "mzr"` option bypasses `Spectra` entirely and streams
batches of spectra directly from `mzR::openMSfile()` into Apache
Arrow's incremental `ParquetFileWriter`:

```{r, eval = FALSE}
be <- mzMLToParquet("huge-sample.mzML",
                    path = tempfile(),
                    engine = "mzr",
                    batch_size = 1000L)
```

Memory use stays bounded by `batch_size * mean(peaks per spectrum)`,
regardless of total file size. The streaming engine processes files
serially and requires the `mzR` package; it does not honour `BPPARAM`
or `chunksize`.

# Saving and reloading the backend

Because `MsBackendParquet` objects only contain a file system path,
they are fully serialisable and can be sent to parallel workers
without re-opening connections:

```{r save-load}
rds <- tempfile(fileext = ".rds")
saveRDS(be, rds)
be2 <- readRDS(rds)
identical(spectraNames(be), spectraNames(be2))
```

# Session information

```{r si}
sessionInfo()
```
