---
title: "Apache Parquet-based Mass Spectrometry Data Backend"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Apache Parquet-based Mass Spectrometry Data Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{MsBackendParquet}
---

```{r style, echo = FALSE, results = "asis", message = FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendParquet")`<br />
**Authors**: `r packageDescription("MsBackendParquet")[["Author"]] `<br />
**Compiled**: `r date()`

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BiocStyle)
```

# Introduction

The `r Biocpkg("MsBackendParquet")` package provides a backend for
`r Biocpkg("Spectra")` objects that stores mass spectrometry (MS) data
in an on-disk **mzStack** dataset, built on
[Apache Parquet](https://parquet.apache.org/).
Datasets are *written* with the
[Apache Arrow](https://arrow.apache.org/docs/r) R bindings and *queried*
with [DuckDB](https://duckdb.org/), keeping the spectra metadata
together with the `mz`/`intensity` peak values in a columnar format that
supports compression, dictionary encoding and Hive-style partitioning.

An mzStack dataset is a directory identified by an `mzStack.json`
manifest, holding one or more *runs*. A run is either `native` — converted
from raw MS data files, with peaks stored alongside the metadata — or
`mzpeak`, an index over external
[HUPO-PSI mzPeak](https://github.com/HUPO-PSI/mzPeak) archives that are
read where they are and never modified (see
*Querying collections of mzPeak archives* below). The manifest records
which kind each run is, so the two never have to be told apart by
inspecting the directory.

To keep access fast, the (small) non-peak spectra variables are held in
a bounded in-memory index while the peak data is read from disk on
demand. As a result, metadata filters cost about as much as with an
in-memory backend, while the bulk peak storage stays out of core.

The backend extends `MsBackendCached` from `r Biocpkg("Spectra")` so it
benefits from caching of user-defined spectra variables while keeping
the on-disk dataset read-only.

# Creating a Parquet dataset

The quickest way to turn raw MS data files (mzML, mzXML, netCDF) into a
Parquet dataset is the `mzMLToParquet()` helper. It validates the input
files, writes the dataset and returns an initialised backend ready to be
wrapped in a `Spectra` object:

```{r, eval = FALSE}
library(Spectra)
library(MsBackendParquet)

files <- c("sample-1.mzML", "sample-2.mzML")
be <- mzMLToParquet(files, path = tempfile(), partitioning = "dataOrigin")
sps <- Spectra(be)
```

For more control (a custom reader backend, chunk sizes, compression
codec, Parquet row-group size, or writing an existing `DataFrame`), call
`createMsBackendParquetDataset()` directly:

```{r, eval = FALSE}
path <- tempfile()
createMsBackendParquetDataset(path = path, x = files,
                              partitioning = "dataOrigin",
                              compression = "snappy")
be <- backendInitialize(MsBackendParquet(), path = path)
```

Both functions import MS data via `MsBackendMzR()` by default and write
spectra metadata together with the `mz`/`intensity` values as Parquet
`list<double>` columns.

For the rest of the vignette we work with a small in-memory data set
so the document can be built without external files:

```{r build-dataset}
library(Spectra)
library(MsBackendParquet)

sd <- S4Vectors::DataFrame(
    msLevel = c(1L, 1L, 2L),
    rtime = c(1.0, 2.0, 3.0),
    precursorMz = c(NA_real_, NA_real_, 110.5),
    dataOrigin = c("file-a", "file-a", "file-b"))
sd$mz <- IRanges::NumericList(c(100, 110), c(101, 111),
                              c(102, 112), compress = FALSE)
sd$intensity <- IRanges::NumericList(c(10, 20), c(11, 21),
                                     c(12, 22), compress = FALSE)

path <- tempfile()
be <- backendInitialize(MsBackendParquet(), path = path, data = sd)
be
```

# Accessing data

`spectraData()`, `peaksData()`, `mz()`, `intensity()` and related work
the same way as for any other `MsBackend`:

```{r access}
spectraData(be, c("msLevel", "rtime"))
peaksData(be)[1:2]
mz(be)
```

# Filtering with DuckDB predicate push-down

Filter methods are translated into DuckDB SQL predicates that are pushed
down to the Parquet scan. DuckDB prunes whole row groups using their
min/max statistics (and skips whole partitions for Hive-partitioned
datasets), so a range or set filter reads only the data it needs. When
the filter column is held in the in-memory metadata index (see below),
the predicate is evaluated in memory instead:

```{r filtering}
filterMsLevel(be, 1L)
filterRt(be, c(1.5, 2.5))
filterDataOrigin(be, "file-b")
filterPrecursorMzRange(be, c(110, 111))
filterPrecursorMzValues(be, 110.5, ppm = 20)
```

Chained filters combine into a single scan, and a subsequent
`peaksData()` call reuses the same predicate to prune the peak data it
reads.

# Querying with MassQL

The `r Biocpkg("SpectraQL")` package adds support for the [Mass Spec Query
Language](https://www.massql.org/) (MassQL) to `Spectra` objects. Because
`MsBackendParquet` is an ordinary `MsBackend`, `query()` works on it
unchanged, and the conditions of a MassQL expression are translated into
exactly the `Spectra` filter calls that the previous section pushes down
into DuckDB.

This section works on a real DDA data file from `r Biocpkg("MsDataHub")`. The
file is downloaded once and cached locally; converting it to Parquet takes
a few seconds:

```{r spectraql-setup, message = FALSE}
library(SpectraQL)
library(MsDataHub)

fl <- MsDataHub::PestMix1_DDA.mzML()
sps <- Spectra(mzMLToParquet(fl, path = tempfile()))
sps
```

A MassQL expression has the form `"QUERY <type of data> WHERE <condition>
AND <condition>"`. The special data type `*` returns the selected spectra
as a `Spectra` object, so a query reads as a filter. Retention times are
given in *seconds*:

```{r ql-rt}
query(sps, "QUERY * WHERE RTMIN = 200 AND RTMAX = 300")
```

`SpectraQL` translates `RTMIN`/`RTMAX` into a `filterRt()` call, which
`MsBackendParquet` answers from its in-memory metadata index (and records
as a DuckDB predicate for any later peak access). Conditions combine with
`AND`, and queries are case insensitive:

```{r ql-polarity}
query(sps, "query * where rtmin = 200 and rtmax = 300 and polarity = positive")
```

Spectra can also be selected by acquisition number with `SCANMIN`/`SCANMAX`:

```{r ql-scan}
query(sps, "QUERY * WHERE SCANMIN = 100 AND SCANMAX = 200")
```

## Selecting on precursor m/z

`MS2PREC` selects MS2 spectra by precursor m/z. Mass accuracy is given with
the optional `TOLERANCEPPM` and `TOLERANCEMZ` fields, which default to `0`:

```{r ql-prec}
query(sps, "QUERY * WHERE MS2PREC = 278.093:TOLERANCEPPM=30")
```

Several values can be combined with `OR`. This becomes a single
`filterPrecursorMzValues()` call, i.e. one OR-ed predicate and one scan
rather than one scan per value:

```{r ql-prec-or}
query(sps, "QUERY * WHERE MS2PREC = (278.093 OR 124.086):TOLERANCEMZ=0.01")
```

Combining a retention time window with a precursor condition narrows the
selection to the few MS2 spectra of that precursor acquired in the window:

```{r ql-combined}
query(sps, paste("QUERY * WHERE RTMIN = 200 AND RTMAX = 300",
                 "AND MS2PREC = 278.093:TOLERANCEPPM=30"))
```

## Selecting on peaks

The conditions above only look at spectra variables. `MS1MZ`, `MS2PROD`
(equivalently `MS2MZ`) and `MS2NL` instead inspect the peaks themselves, so
these queries read peak data from the Parquet dataset.

`MS1MZ` keeps MS1 spectra that contain a peak with the given m/z:

```{r ql-ms1mz}
query(sps, "QUERY * WHERE MS1MZ = 219.026:TOLERANCEMZ=0.01")
```

`MS2PROD` does the same for fragment peaks of MS2 spectra, and pairs
naturally with a precursor condition:

```{r ql-ms2prod}
query(sps, "QUERY * WHERE MS2PROD = 164.054:TOLERANCEMZ=0.01")

query(sps, paste("QUERY * WHERE MS2PREC = 278.093:TOLERANCEPPM=30",
                 "AND MS2PROD = 164.054:TOLERANCEMZ=0.01"))
```

`MS2NL` looks for a neutral loss from the precursor, i.e. a fragment at
`precursorMz - <value>`:

```{r ql-ms2nl}
query(sps, "QUERY * WHERE MS2NL = 114.041:TOLERANCEMZ=0.01")
```

## Choosing what is returned

The *type of data* part of the query decides what comes back. Besides `*`,
`MS1DATA` and `MS2DATA` return the peak matrices of the selected MS1 or MS2
spectra, `scaninfo(...)` returns their `spectraData()` and `scansum(...)`
their total ion current:

```{r ql-what}
pks <- query(sps, "QUERY MS1DATA WHERE RTMIN = 200 AND RTMAX = 300")
length(pks)
head(pks[[1L]], 3)

sdta <- query(sps, "QUERY scaninfo(MS1DATA) WHERE RTMIN = 200 AND RTMAX = 300")
sdta[1:3, c("msLevel", "rtime", "peaksCount", "totIonCurrent")]

tic <- query(sps, "QUERY scansum(MS1DATA) WHERE RTMIN = 200 AND RTMAX = 300")
head(tic)
```

## What each query costs

The two groups of conditions behave very differently on this backend:

- `RTMIN`/`RTMAX` and `MS2PREC` map to `filterRt()` and
  `filterPrecursorMzValues()`. Both are served from the in-memory metadata
  index, or pushed into the Parquet scan as a DuckDB predicate. No peak
  data is touched at all.
- `POLARITY` and `SCANMIN`/`SCANMAX` are resolved from spectra variables,
  which are equally cheap.
- `MS1MZ`, `MS2PROD`, `MS2NL`, and any result of type `MS1DATA`,
  `MS2DATA` or `scansum(...)` need the peaks. Here the columnar layout
  pays off: DuckDB reads only the `mz`/`intensity` columns, and only for
  the rows that survived the other conditions.

The timings for this query suite across the on-disk backends
are summarised in the *Performance* section below.

# The in-memory metadata cache

The non-peak spectra variables are typically small compared to the peak
data, so `MsBackendParquet` keeps them in a bounded, read-through
in-memory index. The cache is held at the package level, keyed by
dataset path and shared by every backend over that dataset (so it is
independent of the individual, serialisable backend objects). It is
warmed on `backendInitialize()` and dropped automatically whenever the
dataset on disk is (re-)written. It is entirely separate from the
user-modified variables tracked by `MsBackendCached`, which always take
precedence.

The cache is controlled through three options (shown with their
defaults):

```{r, eval = FALSE}
# Enable ("auto", the default) or disable ("off") the metadata cache.
options(MsBackendParquet.cacheMetadata = "auto")

# Columns held by default (the ones the optimised filters need).
options(MsBackendParquet.cacheColumns =
            c("msLevel", "rtime", "precursorMz", "dataOrigin"))

# Budget in cells (rows x columns); ~5e7 is roughly 400 MB of doubles.
# Requests over budget quietly fall back to reading from disk.
options(MsBackendParquet.cacheMaxCells = 5e7)
```

Two further options tune the underlying DuckDB session:
`MsBackendParquet.threads` (maps to DuckDB's `threads`) and
`MsBackendParquet.memoryLimit` (maps to `memory_limit`). Both default to
DuckDB's own defaults when unset.

# Performance

Storing the peaks as columnar `list<double>` values lets DuckDB skip
whole row groups when a query touches only part of the data, while
projection and predicate push-down keep both the columns and rows read
to a minimum. Combined with the in-memory metadata index, this makes
metadata filters roughly as fast as an in-memory backend while the peak
storage stays on disk.

Across the SpectraQL query suite over 380,100 spectra (50 mzML files),
the Parquet backend is the fastest of the on-disk `Spectra` backends on
5 of 7 queries (medians in ms, lower is better):

| query | parquet | mzR | HDF5 | SQL |
|---|---|---|---|---|
| `RTMIN`/`RTMAX` range | **12.24** | 15.44 | 24.05 | 380.19 |
| narrow RT range | 11.23 | **5.33** | 14.76 | 348.17 |
| precursor m/z ± ppm | **14.09** | 21.66 | 28.66 | 336.33 |
| RT and precursor | **12.56** | 17.25 | 27.75 | 389.22 |
| MS1 peak data | **314.82** | 1510 | 1440 | 919.04 |
| MS1 TIC | **506.79** | 1390 | 1490 | 890.82 |
| scan info | 26.09 | **17.27** | 22.30 | 718.91 |

The benchmark scripts live in
[`inst/benchmarks/`](https://github.com/computational-metabolomics/MsBackendParquet/tree/main/inst/benchmarks);
see `inst/benchmarks/performance.md` for the full numbers, the
configuration sweeps over the tunable options, and the design
considerations behind them.

# Querying collections of mzPeak archives

[mzPeak](https://github.com/HUPO-PSI/mzPeak) is the HUPO-PSI successor to
mzML. An mzPeak *archive* holds **one MS run** as a folder (or an
uncompressed ZIP) of Parquet files plus a JSON manifest naming them. That
shape is excellent for distributing and archiving a run, and awkward when
the question spans a thousand runs at once — answering it would mean opening
a thousand archives and parsing a thousand manifests.

`MsBackendParquet` bridges the two. `createMzPeakDataset()` registers a set
of archives as one dataset:

```{r, eval = FALSE}
library(Spectra)
library(MsBackendParquet)

path <- tempfile()
createMzPeakDataset(c("QC01.mzpeak", "QC02.mzpeak"), path = path)

be <- backendInitialize(MsBackendParquet(), path = path)
sps <- Spectra(be)
```

**The archives are neither modified nor copied.** Ingest reads each
archive's manifest, its spectrum metadata table and its scan / precursor /
selected-ion side tables, and writes a small derived index next to them —
around 3% of the size of the archives. It never reads a byte of signal data,
so the cost scales with the number of spectra rather than the size of the
data. Peaks are then read from the archives themselves, on demand.

Because the archives stay exactly as they were, they remain readable by
every other mzPeak tool, and exporting a run back out of the dataset is a
directory copy rather than a conversion.

## What the dataset does with mzPeak's model

A few things are translated on the way in, and it is worth knowing which:

- **Units and names.** mzPeak stores retention time as `time` in *minutes*;
  `Spectra` expects `rtime` in *seconds*. The conversion, and every other
  column rename and re-encoding (`scan_polarity` ±1, `spectrum_representation`
  as a CURIE, the isolation window as a target plus offsets, ...), happens in
  a single SQL view. The same view serves a natively converted dataset: its
  `spectra/` files are written with mzPeak's column names and encodings too,
  so both kinds are directly comparable and either can be exported back to an
  archive mechanically. `spectrum_index` is always the dataset's unique
  0-based key; a spectrum's position within its source file (`scanIndex`,
  which restarts per file) is kept in `scan_index`. The package adds only a
  handful of reserved columns (`spectrum_id_`, `acquisition_num_`, the `n_*`
  counts).
- **Flattening.** mzPeak keeps scans, precursors and selected ions in
  separate tables linked back to the spectrum by a foreign key. The index
  takes the first of each, and records how many there were in `n_scans`,
  `n_selected_ions` and `n_precursors`. Nothing is lost — the archive still
  holds the full detail — but a spectrum with several scans is summarised in
  the index.
- **Profile and centroid.** mzPeak can hold both representations of the same
  spectrum, in separate files. Choose with `representation`:

```{r, eval = FALSE}
be <- backendInitialize(MsBackendParquet(), path = path,
                        representation = "centroid")
```

- **Differing schemas.** Two perfectly valid archives often have different
  columns, because writers are encouraged to promote whichever parameters
  they have into columns. They still combine into one queryable dataset;
  columns missing from an archive read as `NA`.

Archives that use mzPeak's *chunked* signal layout are rejected with a clear
message; only the *point* layout is supported so far.

## Searching for peaks across runs

Filters on spectra variables work as they do for any other dataset. Finding
spectra that *contain a peak* at a given m/z is different: it has to look at
the signal, and mzPeak stores signal ordered by spectrum, which means a
filter on m/z cannot skip anything (every block of rows spans the whole mass
range).

`filterContainsMz()` runs that search inside the storage engine rather than
pulling peaks into R:

```{r, eval = FALSE}
hits <- filterContainsMz(be, 278.093, ppm = 20)
```

For datasets where this is a common query, build an m/z-ordered copy of the
signal. Sorting is what makes the block statistics useful, so the same
search then reads a handful of blocks instead of everything:

```{r, eval = FALSE}
buildProjection(path)          # optional, per-run, deletable
```

On a 5-million-point test dataset this made `filterContainsMz()` around eight
times faster. A projection is a **cache**: it is a second copy of the signal
and so costs roughly as much disk again, it can be dropped with
`dropProjection()`, and results are identical with or without it — only the
speed changes. Build it per run, for the runs you query often.

# Streaming large mzML files

For very large mzML files, the default conversion path goes through a
`Spectra` object built with `MsBackendMzR()`, which materialises a
whole chunk of peaks in memory before writing it to Parquet. The
`engine = "mzr"` option bypasses `Spectra` entirely and streams
batches of spectra directly from `mzR::openMSfile()` into Apache
Arrow's incremental `ParquetFileWriter`:

```{r, eval = FALSE}
be <- mzMLToParquet("huge-sample.mzML",
                    path = tempfile(),
                    engine = "mzr",
                    batch_size = 1000L)
```

Memory use stays bounded by `batch_size * mean(peaks per spectrum)`,
regardless of total file size. The streaming engine processes files
serially and requires the `mzR` package; it does not honour `BPPARAM`
or `chunksize`.

# Saving and reloading the backend

Because `MsBackendParquet` objects only contain a file system path,
they are fully serialisable and can be sent to parallel workers
without re-opening connections:

```{r save-load}
rds <- tempfile(fileext = ".rds")
saveRDS(be, rds)
be2 <- readRDS(rds)
identical(spectraNames(be), spectraNames(be2))
```

# Session information

```{r si}
sessionInfo()
```
