---
title: "Performance evaluation and comparison against MsBackendSql"
output:
  html_document: default
---

# Introduction

Compare performance and properties of `MsBackendParquet` compared to
`MsBackendMzR` or `MsBackendSql` on a large data set. This comparison evaluates
benefits of the individual backends when analyzing large-scale mass spectrometry
data. If not otherwise stated, all tests are performed without parallel
processing on a single CPU core.

## MTBLS93

The data set: a subset of 1000 files from MTBLS93 MetaboLights data set.

### Preparing and loading the data

Loading the reference data and creating the data storage modes for the SQLite
and the Parquet-based formats. The code chunks to create the `MsBackendSql` and
`MsBackendParquet`-based formats need only to be run once. After the first
code evaluation the prepared data sets are loaded.

```{r, message = FALSE}
library(Spectra)
library(MsBackendMetaboLights)
register(SerialParam())

sps <- Spectra(mtblsId = "MTBLS93", filePattern = "01.CDF$",
               source = MsBackendMetaboLights())
sps1000_mzr <- filterDataOrigin(sps, unique(dataOrigin(sps))[1:1000])
```

Define the folder where to store the SQLite and Parquet files.

```{r}
## Set the data storage path.
local_path <- "~/tmp/"
sqlite_file <- file.path(local_path, "sps1000.sqlite")
parquet_file <- file.path(local_path, "sps1000.parquet")
```

Create a `MsBackendSql` version of the data. These code chunks

```{r, eval = !file.exists(sqlite_file), message = FALSE}
library(RSQLite)
library(MsBackendSql)
con <- dbConnect(SQLite(), sqlite_file)
system.time(
    sps1000_sql <- setBackend(sps1000_mzr, MsBackendSql(), dbcon = con)
)
```

Create a `MsBackendParquet` version of the data.

```{r, eval = !file.exists(parquet_file), message = FALSE}
library(MsBackendParquet)
system.time(
    sps1000_prq <- setBackend(sps1000_mzr, MsBackendParquet(),
                              path = parquet_file)
)
```

We next load the data also through the `MsBackendOfflineSql` and
`MsBackendParquet` backends.

```{r}
library(RSQLite)
library(MsBackendSql)
library(MsBackendParquet)

sps1000_sql <- Spectra(sqlite_file,
                       source = MsBackendOfflineSql(), drv = SQLite())
sps1000_prq <- Spectra(parquet_file, source = MsBackendParquet())
```

#### Compare sizes

We next compare the object and file sizes for the various backends.

```{r}
print(object.size(sps1000_mzr), units = "MB")
print(object.size(sps1000_sql), units = "MB")
print(object.size(sps1000_prq), units = "MB")
```

Both the `MsBackendSql` and `MsBackendParquet` have similar object size, while
the `MsBackendMzR` is considerably larger. We next compare the file sizes on
disk:

```{r, message = FALSE}
library(fs)
file_size(sqlite_file)

sum(file_size(dir(
    parquet_file, recursive = TRUE, full.names = TRUE)))
```

The `MsBackendParquet` uses about half of the disk space `MsBackendSql`
requires.

#### Extract spectra data

We compare the timing to extract the **full** `spectraData()`. We disable
parallel processing.

```{r}
register(SerialParam())
library(microbenchmark)

microbenchmark(
    a <- spectraData(sps1000_mzr),
    b <- spectraData(sps1000_sql),
    c <- spectraData(sps1000_prq),
    times = 7
)
## Compare object sizes...
print(object.size(sps1000_sql), units = "MB")
print(object.size(sps1000_prq), units = "MB")
```

`MsBackendParquet` is considerably (about 2-3 times) faster than
`MsBackendSql`. Because the full data is already in memory `MsBackendMzR`
outperforms both.

We next compare the performance to extract selected, commonly used, spectra
variables.

```{r}
microbenchmark(
    a <- sps1000_mzr$rtime,
    b <- sps1000_sql$rtime,
    c <- sps1000_prq$rtime,
    times = 7
)

library(testthat)
expect_equal(a, b)
expect_equal(b, c)
```

`MsBackendParquet` is considerably faster than the `MsBackendSql` backend.

#### Extract peaks data

Extracting the peaks data from the full data set would exceed the computers main
memory. We have thus to extract the peaks data from a subset of 50 files
instead (with 100 we also hit the limit of over 96GB main memory).

```{r}
sps50_mzr <- filterDataOrigin(
    sps1000_mzr, unique(dataOrigin(sps1000_mzr))[1:50])
sps50_sql <- filterDataOrigin(
    sps1000_sql, unique(dataOrigin(sps1000_sql))[1:50])
sps50_prq <- filterDataOrigin(
    sps1000_prq, unique(dataOrigin(sps1000_prq))[1:50])

library(peakRAM)
peakRAM(
    peaksData(sps50_mzr),
    peaksData(sps50_sql),
    peaksData(sps50_prq)
)
```

`MsBackendParquet` shows the highest performance followed by
`MsBackendMzR`. `MsBackendSql` is considerably slower.

#### Extract spectra data a small random subset

We restrict the data sets to a small set of random spectra (in arbitrary order).

```{r}
set.seed(123)
idx <- sample(seq_along(sps1000_mzr), 100)

s_mzr <- sps1000_mzr[idx]
s_sql <- sps1000_sql[idx]
s_prq <- sps1000_prq[idx]

microbenchmark(
    spectraData(s_mzr),
    spectraData(s_sql),
    spectraData(s_prq),
    times = 7
)
```

For these small random subsets `MsBackendMzR` shows the highest performance
followed by `MsBackendSql` while `MsBackendParquet` is considerably (about 100
times) slower.

We repeat the same comparison for a single spectra variable.

```{r}
microbenchmark(
    a <- s_mzr$rtime,
    b <- s_sql$rtime,
    c <- s_prq$rtime,
    times = 7
)
expect_equal(a, b)
expect_equal(b, c)
```

For the single spectra variable the performances are somewhat comparable, with
`MsBackendSql` showing the lowest performance.

#### Extract peaks data for a small random subset

We next compare the performance extracting peaks data from the random subsets.

```{r}
microbenchmark(
    a <- peaksData(s_mzr),
    b <- peaksData(s_sql),
    c <- peaksData(s_prq),
    times = 7
)
expect_equal(a, b)
expect_equal(b, c)
```

There are huge differences in the performances of the different
backends. `MsBackendSql` is by far the fastest backend.


#### Extract spectra data for a retention time subset

We repeat the comparison for a second, more realistic, subset: all spectra
within a small retention time range.

```{r}
s_mzr <- filterRt(sps1000_mzr, c(3, 4))
s_sql <- filterRt(sps1000_sql, c(3, 4))
s_prq <- filterRt(sps1000_prq, c(3, 4))
```

```{r}
microbenchmark(
    spectraData(s_mzr),
    spectraData(s_sql),
    spectraData(s_prq),
    times = 7
)
```

`MsBackendMzR` and `MsBackendSql` outperform the `MsBackendParquet` for this
operation. We repeat the comparison for a single spectra variable.

```{r}
microbenchmark(
    a <- s_mzr$rtime,
    b <- s_sql$rtime,
    c <- s_prq$rtime,
    times = 7
)
expect_equal(a, b)
expect_equal(b, c)
```

For the single spectra variable the `MsBackendSql` performs less well.

#### Extract peaks data for a retention time subset

We next compare the performance extracting peaks data from the subsets.

```{r}
peakRAM(
    a <- peaksData(s_mzr),
    b <- peaksData(s_sql),
    c <- peaksData(s_prq)
)
expect_equal(a, b)
expect_equal(b, c)
```

The `MsBackendSql` shows the highest performance, about 3 times faster than the
`MsBackendParquet`. The `MsBackendMzR` is by far the slowest due to the cost of
opening and reading each of the 1000 netCDF files.

#### Extract spectra data for a single data file of a large dataset

Finally, we evaluate performance on a subset of the large data representing
spectra from the (same) originating file.

```{r}
s_mzr <- filterDataOrigin(sps1000_mzr, unique(dataOrigin(sps1000_mzr))[13L])
s_sql <- filterDataOrigin(sps1000_sql, unique(dataOrigin(sps1000_sql))[13L])
s_prq <- filterDataOrigin(sps1000_prq, unique(dataOrigin(sps1000_prq))[13L])
```

```{r}
microbenchmark(
    spectraData(s_mzr),
    spectraData(s_sql),
    spectraData(s_prq),
    times = 7
)
```

Similar to the other data subsets, `MsBackendParquet` is slowest when the full
spectra data is extracted, even if only for one data file.

```{r}
microbenchmark(
    a <- s_mzr$rtime,
    b <- s_sql$rtime,
    c <- s_prq$rtime,
    times = 7
)
expect_equal(a, b)
expect_equal(b, c)
```

In contrast, extracting a single column/spectra variable is very fast with
`MsBackendParquet`, while `MsBackendSql` takes about 4 times longer.

#### Extract peaks data for a single data file of a large dataset

We next compare the performance extracting peaks data from the subsets.

```{r}
microbenchmark(
    a <- peaksData(s_mzr),
    b <- peaksData(s_sql),
    c <- peaksData(s_prq),
    times = 7
)
expect_equal(a, b)
expect_equal(b, c)
```

The performances of `MsBackendMzR` and `MsBackendParquet` are comparable, while
`MsBackendSql` is about 8 times faster than both.

## Summary

- `MsBackendParquet` shows higher performance extracting single spectra
  variables or also peaks data from a large data set or large subsets of
  that data.
- `MsBackendParquet` uses about half of the disk space required by
  `MsBackendSql`.
- The performance of `MsBackendSql` is higher when data from a small (or
  unordered) subset of data is requested (or from a single file of a larger data
  set).
- `MsBackendSql` stores all data in a single database, thus, extracting random
  data from that database is faster because it avoids opening and accessing data
  from different files.

# Session information

```{r}
sessionInfo()
```
